Step by Step: Using Step Selection Functions to Simulate Dispersal and Assess Landscape Connectivity

### 2021-07-16 Chapter_1/04_Manuscript/Manuscript__2021_07_15.tex
Rewriting introduction

\textbf{Keywords:} dispersal, simulation, movement, integrated step selection
function, Kavango-Zambezi Transfrontier Conservation Area, landscape
connectivity, Lycaon pictus

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}

%------------------------------------------------------------------------------
% Short Version
%------------------------------------------------------------------------------
% The ability to disperse is contingent a sufficient degree of landscape
% connectivity, which is why the identification and preservation of movement
% corridors that promote connectivity has become a task of extraordinary
% importance. Currently, ecologists mainly rely on least-cost analysis or circuit
% theory to investigate connectivity, albeit both methods make several assumptions
% are hardly met in reality. To address these issues, simulations from
% individual-based movement models have been proposed, yet a unified framework to
% simulate dispersal and quantify connectivity is lacking.
%
% Here, we employ a simple three-step workflow that combines several
% already-existing methods to assess connectivity using simulated dispersal
% trajectories. In the first step, we use integrated step selection functions to
% parametrize a mechanistic movement model rendering dispersal behavior. In a
% second step, we apply the parametrized model as an individual-based movement
% model to simulate dispersal trajectories. Finally, in the third step, we
% consolidate simulated trajectories into three complementary connectivity maps,
% each focusing on a different aspect of landscape connectivity.
%
% We apply these steps to assess landscape connectivity for dispersing African
% wild dogs (\textit{Lycaon pictus}) southern Africa. We thereby demonstrate that
% simulations from integrated step selection functions offer a simple, yet
% powerful alternative to traditional connectivity modeling techniques.

%------------------------------------------------------------------------------
%  LONG VERSION
%------------------------------------------------------------------------------
% Importance of Dispersal & Connectivity
Dispersal of individuals is a crucial pre-requisite for long-term species
persistence. However, the ability to disperse is contingent a sufficient degree
of landscape connectivity, which is why the identification and preservation of
movement corridors that promote connectivity has become a task of exceptional
importance.

% Quantifying connectivity
Over the past two decades, ecologists have primarily relied on least-cost
analysis and circuit theory to investigate connectivity. Unfortunately, both
methods make several restrictive assumptions that limit their usefulness in
reality. Least-cost analysis, for instance, assumes that animals move towards a
pre-conceived endpoint and choose a cost-minimizing route accordingly. Circuit
theory, in contrast, presumes a random walk without any directional persistence.
To address these shortcomings, simulations from individual-based movement models
have been proposed and applied, yet a unified framework to simulate dispersal
and to derive connectivity maps based on simulated dispersal events is lacking.

% What we do
% and synthesize a simple
% three-step workflow that uses
Here, we combine several already-existing methods to simulate dispersal and
investigate landscape connectivity in three distinct steps. In the first step,
integrated step selection functions are used to parametrize a mechanistic
movement model. In the second step, the parametrized model is applied as an
individual-based movement model to simulate dispersal trajectories. In the third
step, simulated trajectories are consolidated and synthesized into three
complementary connectivity maps, each focusing on a different aspect of
landscape connectivity. The set of maps includes a heatmap, revealing frequently
traversed areas, a betweenness-map, delineating critical dispersal corridors,
and a map of inter-patch connectivity, depicting the presence or absence of
functional links between distinct habitat patches.

% Case study
As a case study, we apply the proposed workflow to assess landscape connectivity
for dispersing African wild dogs (\textit{Lycaon pictus}) in the Kavango-Zambezi
Transfrontier Conservation Area, the world's largest transboundary conservation
area. With this, we demonstrate that the outlined workflow offers a simple, yet
powerful alternative to traditional connectivity modeling techniques. Although
such individual-based simulations necessitate informed decisions about the
number of simulated individuals and the duration of simulated dispersal events,
they enable to relax several of the unrealistic assumptions inherent to
least-cost analysis or circuit theory.

@@

% Create linenumbers
\linenumbers

\section{Introduction}

% Importance of Dispersal & Connectivity
\subsection{Dispersal \& Connectivity Models (90\%)}
Successful dispersal of individuals requires a sufficient degree of landscape
connectivity \citep{Fahrig.2003, Clobert.2012}, which is why the identification
and protection of major dispersal corridors has become a fundamental task in
conservation science \citep{Nathan.2008, Doerr.2011, Rudnick.2012}. To pinpoint
relevant dispersal corridors, a comprehensive understanding of the factors that
limit dispersal of the focal species and an adequate connectivity model are
required \citep{Baguette.2013, Vasudev.2015, Hofmann.2021}. To date, the two
most prominent connectivity models are least-cost path analysis (LCPA;
\citealp{Adriaensen.2003}) and circuit theory (CT; \citealp{McRae.2006,
McRae.2008}), both graph-based approaches that estimate conductance of the
landscape based on predicted habitat permeability \citep{Zeller.2012}.

% Issues with Both Methods
\subsection{Issues with Traditional Methods (90\%)}
Despite their intuitive nature and ease of use, both methods make several
assumptions that are hardly ever met in reality \citep{Diniz.2020}. With LCPA,
for instance, a least costly route always exists, even if associated movement
costs are unreasonably high and will never be incurred by a dispersing
individual. The method also presumes that animals move towards a pre-conceived
endpoint and choose a cost-minimizing route accordingly. While these assumptions
may by fulfilled by migrating animals, dispersers usually move into unfamiliar
territory and are therefore less likely to be aware of associated movement costs
\citep{Koen.2014, Abrahms.2017, Cozzi.2020}. CT relaxes these assumptions, yet
posits that animals move according to a random walk, entailing that directional
biases cannot be rendered \citep{Diniz.2020}. In reality, however, directional
persistence is regularly observed \citep{Bovet.1991, Schultz.2001}, especially
in dispersing individuals \citep{Cozzi.2020, Hofmann.2021}. Some of these
deficiencies can be addressed using alternative graph-based approaches, such as
least-cost \textit{corridors} \citep{Pinto.2009}, \textit{thresholded}
least-cost paths \citep{Landguth.2012}, or \textit{randomized} least-cost paths
\citep{Panzacchi.2016}, yet a certain degree of arbitrariness remains. Finally,
none of the methods is capable of rendering the temporal dimension of dispersal
\citep{Diniz.2020}, meaning that statements about the expected duration required
to traverse a certain corridor are impossible.

% What about IBMMs?
\subsection{What about IBMMs? (90\%)}
To overcome the unrealistic assumptions inherent to traditional connectivity
modeling techniques, simulations from individual-based movement models (IBMMs)
have been proposed and applied \citep{Diniz.2020}. In these models, dispersal
trajectories are simulated based on movement rules that determine how
individuals move across or interact with the prevailing landscape
\citep{Gustafson.1996, Gardner.2004, Graf.2007, KramerSchadt.2004, Revilla.2004,
Revilla.2008, Kanagaraj.2013, Clark.2015, Allen.2016, Hauenstein.2019,
Zeller.2020, Vasudev.2021}. Instead of enforcing connections towards a known
endpoint, IBMMs allow individuals to choose their route as they move and enable
to render potential interactions between movement behavior and environmental
conditions. IBMMs furthermore enable to model directional persistence and to
study the temporal dimension of dispersal \citep{Diniz.2020}. Despite these
advantages, however, a unified framework to simulate dispersal and to derive
connectivity maps based on simulated dispersal events is lacking.

% The Power of Step Selection Functions
\subsection{Step Selection Functions (90\%)}
A promising framework that has been developed more recently...
Over the past decade, the increased availability of large amounts of data and an
increased computational power have led to the development of numerous techniques
to study animal movement and resource selection during dispersal
\citep{Boyce.2002, Fortin.2005, Cushman.2010, Zeller.2012}. Especially
\textit{resource selection functions} \citep{Boyce.2002}, \textit{step selection
functions} \citep{Fortin.2005}, and \textit{path selection functions}
\citep{Cushman.2010} have pushed the field of movement ecology forward. These
methods allow estimating habitat preferences of the focal species by comparing
covariates at locations visited by the animal to the same covariates at
locations available to, but not visited by the animal \citep{Boyce.2002,
Fortin.2005, Cushman.2010, Thurfjell.2014}. While regular SSFs were intended to
learn about relative habitat preferences of the focal species
\citep{Fortin.2005, Thurfjell.2014, Avgar.2017}, the method has recently been
generalized to \textit{integrated} SSFs and now enables to jointly study habitat
and movement preferences, as well as potential interactions between them
\citep{Avgar.2016, Signer.2017, Fieberg.2020}. ISSFs therefore provide a
relatively simple means to model complex movement behavior, where movement is
viewed as the result of two intertwined behavioral kernels (e.g.
\citealp{Prokopenko.2017, Munden.2020}). Importantly, a parametrized ISSF model
represents a fully mechanistic movement model based on which individual movement
trajectories can be simulated \citep{Avgar.2016, Signer.2017}. In fact,
\cite{Signer.2017} used ISSF to simulate steady state utilization distributions
of resident animals that were moving around a point of attraction. However, the
degree to which such simulations are helpful in detecting movement corridors and
modeling landscape connectivity remains to be investigated.

% What we do
\subsection{Proposed Framework (90\%)}
Here, we therefore propose to combine several already-existing methods into a
simple three-step workflow to simulate dispersal and investigate landscape
connectivity. In the first step, integrated step selection functions are used to
parametrize a mechanistic movement model rendering dispersal behavior. In the
second step, the parametrized model is applied as an individual-based movement
model to simulate dispersal trajectories. Finally, in the third step, simulated
trajectories are consolidated into three complementary connectivity maps, each
focusing on a different aspect of landscape connectivity. The set of maps
includes a heatmap, revealing frequently traversed areas, a betweenness-map,
delineating critical dispersal corridors, and a map of inter-patch connectivity,
depicting the presence or absence of functional links between distinct habitat
patches.

% Study Species & Study Area
\subsection{Study Species \& Study Area (90\%)}
We exemplify application of the proposed workflow by investigating landscape
connectivity for the endangered African wild dog (\textit{Lycon pictus}). While
once present across entire sub-Saharan Africa, wild dogs have disappeared from a
vast majority of their historic range, mainly due to persecution by humans,
habitat fragmentation and destruction, and deadly diseases
\citep{Woodroffe.2012}. Within the few remaining subpopulations, wild dogs form
cohesive packs comprising 8 to 12 adults and their offspring \cite{McNutt.1995}.
After reaching sexual maturity, male and female offspring form same-sex
coalitions and disperse from their natal pack in search for potential mating
partners and a suitable territory to settle \citep{McNutt.1996, Behr.2020}. New
packs are formed when dispersing coalitions join unrelated opposite-sex
dispersing coalitions \citep{McNutt.1996}. During dispersal, wild dogs can cover
several hundred kilometers across a variety of landscapes
\citep{DaviesMostert.2012, Masenga.2016, Cozzi.2020, Hofmann.2021}. One of the
main strongholds for this species lies near the Moremi Game Reserve in northern
Botswana, which is part of the world's largest transboundary protected area,
namely the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA). This
area was originally devoted towards facilitating the migration of elephants, yet
it can be expected to yield benefits to a great multitude of other species
\citep{Elliot.2014, Brennan.2020, Hofmann.2021}.

% Previous Paper
\subsection{Previous Paper (90\%)}
We previously assessed landscape connectivity for dispersing African wild dogs
within the KAZA-TFCA using least-cost methods \citep{Hofmann.2021}. More
specifically, we fitted a habitat selection model and predicted a permeability
surface based on which we computed least-cost paths and corridors between
protected areas. We now expand on this knowledge and use ISSFs to develop a more
detailed, mechanistic movement model of dispersing wild dogs
(\Cref{GraphicalAbstract}). We then employ the model to simulate 80'000
dispersers moving 2'000 steps across the KAZA-TFCA. Finally, we generate a
heatmap, betweenness map and a map of interpatch connectivity. We thereby
exemplify how ISSFs can be utilized for dispersal simulations and we discuss
several benefits of this approach over traditional connectivity modeling
techniques such as least-cost analysis and circuit theory. Most importantly,
simulations based on ISSFs provide a more generic view on how connectivity
emerges and to which degree connectivity depends on the dispersal duration. In
addition, by generating proper dispersal trajectories, network theory can be
applied to calculate network metrics that are pertinent to connectivity
analysis.

@@

\subsection{GPS Relocation Data (90\%)}
GPS data of 16 dispersing wild dogs was obtained as described in
\citep{Cozzi.2020, Hofmann.2021}. We collected data on 7 female and 9 male
dispersing coalitions between 2011 and 2019 from a free-ranging population
inhabiting the Moremi National Park in northern Botswana. In some instances,
exact dispersal dates were known from field observations, in other cases we
determined dispersal phases using the net-squared displacement metric. Net
squared displacement measures the squared Euclidean distance of a GPS relocation
to a reference point \citep{Borger.2012}, which in our case was set to the
center of each individual's natal home range. Thus, dispersal was deemed to have
started when an individual left its natal home range and ended once individuals
became sedentary again. Because behavior during dispersal is more pertinent to
landscape connectivity \citep{Elliot.2014, Abrahms.2017}, we discarded all data
that was collected during residency. During dispersal, collars were programmed
to record a GPS fix every four hours and regularly transmitted over the Iridium
satellite system. As previous research found no differences in behaviors of
females and males during dispersal \citep{Woodroffe.2019, Cozzi.2020}, we did
not distinguish between sexes. After collection, we converted GPS coordinates (n
= 4'169) to steps, where each step represented the straight-line distance
traveled by an individual between two consecutive GPS relocations
\citep{Turchin.1998}. To ensure a regular sampling interval, we removed fixes
that were not successfully collected on the 4-hourly schedule (\( \pm \) 15
minutes).

\subsection{Covariates (90\%)}
We represented the physical landscape across the study area using a set of
habitat covariates that included water-cover, distance to water, woodland-cover,
and shrub/grassland-cover. Because water cover greatly changes within and
between years in the Okavango Delta, we applied a remote sensing algorithm and
generated frequently updated water cover layers and corresponding distance to
water layers (see \citealp{Wolski.2017} and Appendix A3 in
\citealp{Hofmann.2021}). Resulting layers thus temporally aligned with each
dispersal event, ensuring minimal temporal lag between observed movements and
corresponding water layer. We furthermore computed a proxy for human influence,
rendering anthropogenic pressures stemming from human-density, agricultural
sites, and roads. All spatial layers were coarsened or interpolated to a target
resolution of 250 m by 250 m. A detailed description of the preparation of each
habitat covariate is given in \cite{Hofmann.2021}.

Besides habitat covariates, we computed movement metrics that we used as
movement covariates in our models. Movement metrics were calculated for each
step and included the step length (\textsf{sl}), its natural logarithm
(\textsf{log(sl)}), and the cosine of the relative turning angle
(\textsf{cos(ta)}) (for details see \citep{Avgar.2016, Fieberg.2020}). Because
wild dogs follow a diurnal activity pattern \citep{Castello.2018}, we also coded
a binary variable (\textsf{LowActivity}) indicating whether a step was realized
during periods of low wild dog activity (17:00 to 09:00 local time) or high wild
dog activity (09:00 to 17:00 local time). Handling and manipulation of all data,
as well as all models and simulations were implemented with the statistical
software {\tt R}, version 3.6.6 \citep{R.2019}. Several helper functions were
written in {\tt C++} and imported into {\tt R} using the {\tt Rcpp} package
\citep{Eddelbuettel.2011, Eddelbuettel.2013}.

@@

The structure of the movement model was based on the habitat selection model for
dispersing wild dogs presented in \cite{Hofmann.2021}. In the original model
(hereafter referred to as base model), no interactions between habitat
covariates (\textsf{Water, DistanceToWater\textsuperscript{0.5}, Woodland,
Shrubs/Grassland, Human Influence}) and movement covariates (\textsf{sl,
log(sl), cos(ta)}) were considered. Hence, we slightly expanded this base model
and proposed interactions between all movement and habitat covariates. More
specifically, we started with the base model and incrementally increased model
complexity by adding all possible two-way interactions between habitat
covariates and movement covariates. For instance, for the covariate
\textsf{Water}, we proposed the interactions \textsf{Water:log(sl)},
\textsf{Water:log(sl)}, and \textsf{Water:cos(ta)}. Besides those combinations,
we also proposed the interactions \textsf{sl:cos(ta)} and
\textsf{log(sl):cos(ta)} to account for a correlation between turning angles and
step lengths, as well as the interactions \textsf{sl:LowActivity} and
\textsf{log(sl):LowActivity} to account for the fact that step lengths may
differ due to wild dogs' diurnal activity pattern. To compare competing models
and assess the most parsimonious movement model, we ran stepwise forward model
selection based on Akaike's Information Criterion (AIC, \citealp{Burnham.2002}).

@@

\subsection{Convergence (80\%)}
To verify that the number of simulated individuals sufficed to ensure reliable
estimates of connectivity, we evaluated how the relative traversal frequency
across the landscape changed as we increased the number of simulated
individuals. For this, we distributed 1'000 rectangular ``checkpoints'', each
with an extent of 5 km x 5 km at random locations inside the main study area. We
then determined the relative traversal frequency by simulated trajectories
through each checkpoint as we increased the number of simulated individuals (1
to 50'000). To assess variability in the relative traversal frequency, we
repeatedly sampled trajectories 100 times and computed the mean traversal
frequency across replicates, as well as the 95\% prediction-interval. We deemed
that a checkpoint converged as soon as the width of the prediction interval for
the traversal frequency across replicates dropped below a value of 0.01.

\subsection{Heatmap (100\%)}
To identify dispersal hotspots across our study area, we created a heatmap
indicating the absolute frequency at which each raster-cell in the study area
was visited by virtual dispersers \citep{Hauenstein.2019, Peer.2008}. For this,
we rasterized all simulated trajectories and tallied them into a single map. If
the same trajectory crossed a raster-cell twice, we only counted it once,
thereby mitigating potential biases caused by individuals that moved in circles
because they were trapped by dispersal barriers. To achieve high performance
rasterization, we used the R-package {\tt terra} \citep{Hijmans.2020}.

>>>>

\textbf{Keywords:} dispersal, simulation, movement, integrated step selection,
Kavango-Zambezi Transfrontier Conservation Area, landscape connectivity, Lycaon
pictus

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
% Importance of Dispersal & Connectivity
Dispersal of individuals is a crucial pre-requisite for long-term species
persistence. However, the ability to disperse is contingent a sufficient degree
of landscape connectivity, which is why the identification and preservation of
movement corridors that promote connectivity has become a task of extraordinary
importance for conservation authorities.

% Quantifying connectivity
During the past two decades, ecologists have largely relied on graph-based
methods, such as least-cost analysis or circuit theory, to investigate
connectivity. Despite their popularity, both methods make several restrictive
assumptions that limit their usefulness in reality. Least-cost analysis, for
instance, assumes that animals move towards a pre-conceived endpoint and choose
a cost-minimizing route accordingly. Circuit theory, in contrast, presumes that
animals move according to a random walk without any directional persistence. To
overcome these shortcomings, dispersal simulations from individual-based
movement models have been proposed, yet a unified framework to simulate
dispersal and assess connectivity is lacking.

% What we do
Here, we examine the use of integrated step selection analysis as a simple but
powerful framework for fitting individual-based movement models and simulating
dispersal. For this, we analyse data of 16 dispersing wild dogs (\textit{Lycaon
pictus}) and parametrize a fully mechanistic movement model describing how
dispersers move through the available landscape. We then employ the model to
simulate 80'000 dispersing wild dogs moving across the world's largest
transboundary conservation area. Finally, we use the simulated trajectories to
derive three complementary connectivity maps, each focusing on a different
aspect of connectivity. The set of maps includes a heatmap, revealing frequently
traversed areas, a betweenness-map, delineating critical dispersal corridors,
and a map of inter-patch connectivity, depicting the presence or absence of
functional links between distinct habitat patches.

% Relevance
We conclude that simulations from integrated step selection functions indeed
offer a powerful alternative to traditional connectivity modeling techniques and
requires fewer unrealistic assumptions about dispersal. Nevertheless, some
non-trivial modeling decisions such as the number and duration of simulated
dispersal events need to be considered.

@@

% Create linenumbers
% \linenumbers

\section{Introduction}
% Importance of Dispersal & Connectivity
\subsection{Dispersal \& Connectivity (90\%)}
Successful dispersal of individuals requires a sufficient degree of landscape
connectivity \citep{Fahrig.2003, Clobert.2012}, which is why the identification
and protection of major dispersal corridors has become a fundamental task in
conservation science \citep{Nathan.2008, Doerr.2011, Rudnick.2012}. Pinpointing
relevant dispersal hotspots not only necessitates information on movement
behavior and knowledge about factors that limit dispersal \citep{Baguette.2013,
Vasudev.2015, Hofmann.2021}, but also an adequate connectivity model
\citep{Diniz.2020}. To date, two of the most prominent connectivity models used
to infer movement corridors are least-cost path analysis (LCPA;
\citealp{Adriaensen.2003}) and circuit theory (CT; \citealp{McRae.2006,
McRae.2008}), both graph-based approaches that estimate conductance of the
landscape based on estimated habitat permeability for the focal species
\citep{Zeller.2012}.

% Issues with Least-Cost Paths
\subsection{Issues with LCPs and CT (90\%)}
Despite their intuitive nature and ease of use, least-cost analysis and circuit
theory make several rigorous assumptions that are hardly ever met in reality
\citep{Diniz.2020}. With LCP analysis, for instance, a least costly route always
exists, even if associated movement costs are unreasonably high and will never
be incurred by a dispersing individual. The method also presumes that animals
move towards a pre-conceived endpoint and choose a cost-minimizing route
accordingly, which further requires perfect knowledge about landscape and
associated movement costs. While this may be a reasonable assumption for
migrating animals, dispersers usually move into unfamiliar territory and are
therefore unaware of associated movement costs and potential endpoints
\citep{Koen.2014, Abrahms.2017, Cozzi.2020}. Circuit theory relaxes these
assumptions, yet requires that animals move according to a random walk,
entailing that directional biases cannot be rendered \citep{Diniz.2020}.
Nevertheless, directional movements are regularly observed \citep{Bovet.1991,
Schultz.2001}, particularly in dispersing individuals \citep{Cozzi.2020,
Hofmann.2021}. Some of these deficiencies can be addressed using intermediate
methods, such as (e.g. least-cost \textit{corridors} \citep{Pinto.2009},
\textit{thresholded} least-cost paths \citep{Landguth.2012}, and
\textit{randomized} least-cost paths \citep{Panzacchi.2016}, yet a certain
degree of arbitrariness remains. In addition, neither of these methods is
capable of rendering the temporal dimension of dispersal \citep{Diniz.2020},
such that statements about the expected duration required to traverse a certain
corridor are impossible.

% What about IBMMs?
\subsection{What about IBMMs? (90\%)}
To overcome the issues inherent to LCPs and CT, individual-based movement models
(IBMMs) have been proposed and applied \citep{Diniz.2020}. In these models,
individual dispersal trajectories are simulated based according to movement
rules that determine how individuals move across or interact with the prevailing
landscape \citep{Gustafson.1996, Gardner.2004, Graf.2007, KramerSchadt.2004,
Revilla.2004, Revilla.2008, Kanagaraj.2013, Clark.2015, Allen.2016,
Hauenstein.2019, Zeller.2020, Vasudev.2021}. To quantify landscape connectivity
from simulated trajectories, as set of connectivity metrics, such as inter-patch
connectivity and traversal frequency, to reveal major dispersal corridors
\citep{Kanagaraj.2013, BastilleRousseau.2018, Hauenstein.2019, Zeller.2020} can
be calculated. However, while IBMMS can be employed to overcome any of the
shortcomings intrinsic to LCPs and CT, they are subject to a vast amount of
subjective, non-trivial modeling decisions. Moreover, they can be challenging to
fit and require vast amounts of movement data, ideally collected during
dispersal \citep{Diniz.2020}. Consequently, alternative methods that require
fewer modeling decisions and are straight forward to apply are desirable.

% Step Selection Analysis
\subsection{Step Selection Analysis (90\%)}
Here, we investigate the usefulness of integrated step selection functions
(ISSFs, \citealp{Avgar.2016}), as a relatively simple but powerful IBMM based on
which dispersal can be simulated. While regular SSFs were intended to learn
about relative habitat preferences of the focal species \citep{Fortin.2005,
Thurfjell.2014, Avgar.2017}, the method has recently been generalized to
\textit{integrated} SSFs and now enables to jointly study habitat and movement
preferences, as well as potential interactions between them \citep{Avgar.2016,
Signer.2017, Fieberg.2020}. ISSFs therefore provide a relatively simple means to
model complex movement behavior, where movement is viewed as the result of two
intertwined behavioral kernels (e.g. \citealp{Prokopenko.2017, Munden.2020}).
Importantly, a parametrized ISSF model can be employed as a fully mechanistic
movement model based on which individual movement trajectories can be simulated
\citep{Avgar.2016, Signer.2017}. In fact, \cite{Signer.2017} used ISSF to
simulate steady state utilization distributions of resident animals that were
moving around a point of attraction. However, the degree to which such
simulations are helpful in detecting movement corridors and modeling landscape
connectivity remains to be investigated.

% Study Species & Study Area
\subsection{Study Species \& Study Area (90\%)}
One of the species for which long-term viability relies on sufficient landscape
connectivity is the endangered African wild dog \textit{Lycon pictus}. While
once present across entire sub-Saharan Africa, wild dogs have disappeared from a
vast majority of their historic range due to persecution by humans, habitat
fragmentation and destruction, and deadly diseases \citep{Woodroffe.2012}. As of
today, only 6'000 free-ranging individuals remain in small and spatially
scattered subpopulations \citep{Woodroffe.2012}. Within those subpopulations,
wild dogs form cohesive packs comprising 8 to 12 adults and their offspring
\cite{McNutt.1995}. After reaching sexual maturity, male and female offspring
form same-sex coalitions and disperse from their natal pack in search for
potential mating partners and a suitable territory to settle \citep{McNutt.1996,
Behr.2020}. New packs are formed when dispersing coalitions join unrelated
opposite-sex dispersing coalitions \citep{McNutt.1996}. Dispersing wild dogs can
cover several hundred kilometers across a variety of landscapes
\citep{DaviesMostert.2012, Masenga.2016, Cozzi.2020, Hofmann.2021}. One of the
few strongholds for this species lies near the Moremi Game Reserve in northern
Botswana, which is part of the world's largest transboundary protected area,
namely the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA). This
area has originally been intended to facilitate migration of elephants, but is
expected to provide benefits to a multitude of other species \citep{Elliot.2014,
Brennan.2020, Hofmann.2021}.

% Previous Paper
\subsection{Previous Paper (90\%)}
We previously assessed landscape connectivity for dispersing African wild dogs
within the KAZA-TFCA using least-cost methods \citep{Hofmann.2021}.
Specifically, we fitted a habitat selection model and predicted a permeability
surface that we used to compute least-cost paths and corridors between protected
areas. We now expand on this knowledge and use ISSFs to develop a more detailed,
mechanistic movement model of dispersing wild dogs (\Cref{GraphicalAbstract}).
We employ the model to simulate dispersers moving across the KAZA-TFCA and
generate three distinct connectivity maps, each shedding light onto a different
aspect of connectivity. The set of maps includes a heatmap, revealing frequently
traversed areas, a betweenness-map, delineating critical dispersal corridors and
bottlenecks, and a map of inter-patch connectivity, depicting the presence or
absence of functional links between national parks as well as the average
dispersal duration required to realize those links. With this work, we exemplify
how ISSFs can be utilized for dispersal simulations and we discuss several
benefits of this approach over traditional connectivity modeling techniques such
as least-cost analysis and circuit theory. Most importantly, simulations based
on ISSFs provide a more generic view on how connectivity emerges and to which
degree connectivity depends on the dispersal duration. In addition, by
generating proper dispersal trajectories, network theory can be applied to
calculate network metrics that are pertinent to connectivity analysis.

@@

\subsection{GPS Relocation Data (90\%)}
GPS data of dispersing wild dogs was obtained as described in \citep{Cozzi.2020,
Hofmann.2021}. We collected data on 7 female and 9 male dispersing coalitions
between 2011 and 2019 from a free-ranging population inhabiting the Moremi
National Park in northern Botswana. In some instances, exact dispersal dates
were known from field observations, in other cases we determined dispersal
phases using the net-squared displacement metric. Net squared displacement
measures the squared Euclidean distance of a GPS relocation to a reference point
\citep{Borger.2012}, which in our case was set to the center of each
individual's natal home range. Thus, dispersal was deemed to have started when
an individual left its natal home range and ended once individuals became
sedentary again. Because behavior during dispersal is more pertinent to
landscape connectivity \citep{Elliot.2014, Abrahms.2017}, we discarded all data
that was collected during residency. During dispersal, collars were programmed
to record a GPS fix every four hours and regularly transmitted over the Iridium
satellite system. As previous research found no differences in behaviors of
females and males during dispersal \citep{Woodroffe.2019, Cozzi.2020}, we did
not distinguish between sexes. After collection, we converted GPS coordinates (n
= 4'169) to steps, where each step represented the straight-line distance
traveled by an individual between two consecutive GPS relocations
\citep{Turchin.1998}. To ensure a regular sampling interval, we removed fixes
that were not successfully collected on the 4-hourly schedule (\( \pm \) 15
minutes).

\subsection{Covariates (90\%)}
We represented the physical landscape across the study area using a set of
habitat covariates that included water-cover, distance to water, woodland-cover,
and shrub/grassland-cover. Because water cover greatly changes within and
between years in the Okavango Delta, we applied a remote sensing algorithm and
generated frequently updated water cover layers and corresponding distance to
water layers (see \citealp{Wolski.2017} and Appendix A3 in
\citealp{Hofmann.2021}). Resulting water layers thus temporally aligned with
each dispersal event, ensuring minimal temporal lag between observed movements
and corresponding water layer. We furthermore computed a proxy for human
influence, rendering anthropogenic pressures stemming from human-density,
agricultural sites, and roads. All spatial layers were coarsened or interpolated
to a target resolution of 250 m by 250 m. A detailed description of the
preparation of each habitat covariate is given in \cite{Hofmann.2021}.

Besides habitat covariates, we computed movement metrics that we used as
movement covariates in our models. Movement metrics were calculated for each
step and included the step length (\textsf{sl}), its natural logarithm
(\textsf{log(sl)}), and the cosine of the relative turning angle
(\textsf{cos(ta)}) (for details see \citep{Avgar.2016, Fieberg.2020}). Because
wild dogs follow a diurnal activity pattern \citep{Castello.2018}, we also coded
a binary variable (\textsf{LowActivity}) indicating whether a step was realized
during periods of low wild dog activity (17:00 to 09:00 local time) or high wild
dog activity (09:00 to 17:00 local time). Handling and manipulation of all data,
as well as all models and simulations were implemented with the statistical
software {\tt R}, version 3.6.6 \citep{R.2019}. Several helper functions were
written in {\tt C++} and imported into {\tt R} using the {\tt Rcpp} package
\citep{Eddelbuettel.2011, Eddelbuettel.2013}

@@

The structure of the movement model was based on the habitat selection model for
dispersing wild dogs presented in \cite{Hofmann.2021}. In the original model
(referred to as base model hereafter), no interactions between habitat
covariates (\textsf{Water, DistanceToWater\textsuperscript{0.5}, Woodland,
Shrubs/Grassland, Human Influence}) and movement covariates (\textsf{sl,
log(sl), cos(ta)}) were considered. Hence, we slightly expanded this base model
and proposed interactions between all movement and habitat covariates. More
specifically, we started with the base model and incrementally increased model
complexity by adding all possible two-way interactions between habitat
covariates and movement covariates. For instance, for the covariate
\textsf{Water}, we proposed the interactions \textsf{Water:log(sl)},
\textsf{Water:log(sl)}, and \textsf{Water:cos(ta)}. Besides those combinations,
we also proposed the interactions \textsf{sl:cos(ta)} and
\textsf{log(sl):cos(ta)} to account for a correlation between turning angles and
step lengths, as well as the interactions \textsf{sl:LowActivity} and
\textsf{log(sl):LowActivity} to account for the fact that step lengths may
differ due to wild dogs' diurnal activity pattern. To compare competing models
and assess the most parsimonious movement model, we ran stepwise forward model
selection based on Akaike's Information Criterion (AIC, \citealp{Burnham.2002}).

@@

\subsection{Convergence (80\%)}
To verify that the number of simulated individuals sufficed to ensure reliable
estimates of connectivity, we evaluated how the relative traversal frequency
across the landscape changed as we increased the number of simulated
individuals. For this, we distributed 1'000 rectangular ``checkpoints'', each
with an extent of 5 km x 5 km at random locations inside the main study area. We
then determined the relative traversal frequency by simulated trajectories
through each checkpoint for different numbers of simulations (1 to 50'000). To
assess variability in the relative traversal frequency, we repeatedly sampled
trajectories 100 times and computed the mean traversal frequency across
replicates, as well as the 95\% prediction-interval. We deemed that a checkpoint
converged as soon as the width of the prediction interval for the traversal
frequency across replicates dropped below a value of 0.01.

\subsection{Heatmap (100\%)}
To identify dispersal hotspots across our study area, we created a heatmap
indicating the absolute frequency at which each raster-cell in the study area
was visited by virtual dispersers \citep{Hauenstein.2019, Peer.2008}. For this,
we rasterized all simulated trajectories and tallied them into a single map. If
the same trajectory crossed a raster-cell twice, we only counted it once,
thereby mitigating potential biases caused by individuals that were trapped and
moved in circles. To achieve high performance rasterization, we used the
R-package {\tt terra} \citep{Hijmans.2020}.
